Dynamics of Laterally Propagating Flames in X-ray Bursts. I. Burning Front Structure

### 2020-02-18 xrb/submit/paper.tex
update initial model

The temperature profile in the star and ramp region is given as:
\begin{equation}
T(y) = T_\star + \frac{1}{2} (T_\mathrm{hi} - T_\star) \left [ 1 + \tanh\left( \frac{\tilde{y}}{2 \delta_\mathrm{atm}} \right ) \right ]
\end{equation}
with
\begin{equation}
\tilde{y} = y - H_\star - \frac{3}{2} \delta_\mathrm{atm}
\end{equation}
Here, $\delta_\mathrm{atm}$ is a characteristic width of the
transition ramp, $T_\star$ is the temperature of the neutron star, and
$T_\mathrm{hi}$ is the highest temperature in the HSE model---it will
represent the base of the fuel layer. \added{We also define a floor temperature,
$T_\mathrm{lo}$, which the atmosphere reaches and is held constant at in
the outer, isothermal layer.}

@@

We specify the density, $\rho_\mathrm{int} = \rho(y = H_\star)$ as the
starting point for the integration of hydrostatic equilibrium.  This
is just below the ramp-up region---this ensures that regardless of
what the peak temperature ($T_\mathrm{hi}$) is, the state beneath the
ramp-up region remains unchanged.  Therefore, we will still be in
lateral equilibrium in the star region.  We will denote the density
where $T = T_\mathrm{hi}$ as $\rho_\mathrm{fuel}$.

Creating the model involves specifying $T_\star$, $T_\mathrm{hi}$,
$T_\mathrm{lo}$, $\rho_\mathrm{int}$, $H_\star$,
$\delta_\mathrm{atm}$, ${X_k}_\star$, ${X_k}_\mathrm{atm}$, and $g$.  We
then integrate outwards from the base of the ramp region ($y =
H_\star$), enforcing the discrete form of hydrostatic equilibrium,
Eq.~\ref{eq:hse}.  Integrating upwards, we find $p_i$ and
$\rho_i$ using a Newton-Raphson solver together with the equation of state, with either the temperature
specified, $T_i = T(p_i, \rho_i, \{X_k\}_i)$ (in the isothermal, ramp,
and buffer layers) or constant entropy, $s_i = s(p_i, \rho_i,
\{X_k\}_i)$, in the fuel layer.  This follows the procedures described in \citet{ppm-hse}.  We
use a constant temperature for all $y < H_\star +
3\delta_\mathrm{atm}$.  Above this, we switch to an isentropic atmosphere until the
temperature drops to \deleted{a floor value, }$T_\mathrm{lo}$, at which point we
again keep the temperature constant.  The integration of the
atmosphere continues until the density falls to a low density cutoff,
$\rho_\mathrm{cutoff}$.  The material above this height is taken to
have constant density and temperature.
The choice of factors in front of $\delta_\mathrm{atm}$ were designed
to make sure the peak $T$ is attained at the desired density
of the burning layer.
The parameters we use for
the model generation are listed in Table~\ref{table:params} and the 
initial model profiles are showing in Figure~\ref{fig:initial_models}.

\begin{deluxetable}{lcc}
\tablecaption{\label{table:params} Initial model parameters.}
\tablehead{\colhead{parameter} & \colhead{cool} & \colhead{hot}}
\startdata
$T_\star$       & \multicolumn{2}{c}{$10^8$~K} \\
$T_\mathrm{hi}$ & $2\times 10^8$~K & $1.4\times 10^9$~K \\
$T_\mathrm{lo}$ & \multicolumn{2}{c}{$8\times 10^6$~K} \\
$\rho_\mathrm{int}$ & \multicolumn{2}{c}{$3.43\times 10^6~\gcc$} \\
$\rho_\mathrm{fuel}$\tablenotemark{a} & $2.36\times 10^6~\gcc$ & $1.20\times 10^6~\gcc$ \\
$\rho_\mathrm{cutoff}$ & \multicolumn{2}{c}{$10^{-4}~\gcc$} \\
$H_\star$       & \multicolumn{2}{c}{2000~cm} \\
$\delta_\mathrm{atm}$ & \multicolumn{2}{c}{50~cm} \\
$X_\star(\isotm{Ni}{56})\tablenotemark{b}$ & \multicolumn{2}{c}{1.0} \\
$X_\mathrm{atm}(\isotm{He}{4})\tablenotemark{b}$ & \multicolumn{2}{c}{1.0} \\
$\gb$             & \multicolumn{2}{c}{$-1.5\times 10^{14}~\mathrm{cm~s^{-2}} \hat{\bf y}$} \\
\enddata
%
\tablenotetext{a}{This is not an input parameter, but instead is
  computed during integration.  We list it here for reference.}
%
\tablenotetext{b}{All other species are taken as 0.}
\end{deluxetable}

We blend the hot and cold models laterally to produce the perturbation
needed to initiate a localized flame, with the hot model at the
origin of the axisymmetric geometry.  The blending is done as:
\begin{align}
p(x,y) &= f(x) p_\mathrm{hot}(y) + [1-f(x)] p_\mathrm{cool}(y) \\
\rho(x,y) &= f(x) \rho_\mathrm{hot}(y) + [1-f(x)] \rho_\mathrm{cool}(y) \\
X_k(x,y) &= f(x) {X_k}_\mathrm{hot}(y) + [1-f(x)] {X_k}_\mathrm{cool}(y)
\end{align}
with
\begin{equation}
f(x) = \begin{cases}
     1 & x < x_\mathrm{pert} \\
   1 - \frac{x - x_\mathrm{pert}}{\delta_\mathrm{blend}} & x_\mathrm{pert} \le x \le x_\mathrm{pert} + \delta_\mathrm{blend} \\
     0 & x > x_\mathrm{pert} + \delta_\mathrm{blend}
\end{cases}
\end{equation}
Since the equation of hydrostatic equilibrium is linear and our
blending is a linear combination of two models in hydrostatic
equilibrum, the blended model is also in vertical equilibrium initially.  We choose
$x_\mathrm{pert} = 1.024\times 10^4$~cm and $\delta_\mathrm{blend} = 2048$~cm.  Once
the blended model is constructed, we compute $T(x,y)$ and $(\rho e)(x,y)$
from the equation of state,
\begin{align}
  T(x,y) &= T(\rho(x,y), p(x,y), X_k(x,y)) \\
  (\rho e)(x,y) &= \rho(x, y) \cdot e(\rho(x,y), p(x,y), X_k(x,y)) 
\end{align}

>>>>

The temperature profile in the star and ramp region is given as:
\begin{equation}
T(z) = T_\star + \frac{1}{2} (T_\mathrm{hi} - T_\star) \left [ 1 + \tanh\left( \frac{\tilde{z}}{2 \delta_\mathrm{atm}} \right ) \right ]
\end{equation}
with
\begin{equation}
\tilde{z} = z - H_\star - \frac{3}{2} \delta_\mathrm{atm}
\end{equation}
Here, $\delta_\mathrm{atm}$ is a characteristic width of the
transition ramp, $T_\star$ is the temperature of the neutron star, and
$T_\mathrm{hi}$ is the highest temperature in the HSE model---it will
represent the base of the fuel layer. \added{We also define a floor temperature,
$T_\mathrm{lo}$, which the atmosphere reaches and is held constant at in
the outer, isothermal layer.}

@@

We specify the density, $\rho_\mathrm{int} = \rho(z = H_\star)$ as the
starting point for the integration of hydrostatic equilibrium.  This
is just below the ramp-up region---this ensures that regardless of
what the peak temperature ($T_\mathrm{hi}$) is, the state beneath the
ramp-up region remains unchanged.  Therefore, we will still be in
lateral equilibrium in the star region.  We will denote the density
where $T = T_\mathrm{hi}$ as $\rho_\mathrm{fuel}$.

Creating the model involves specifying $T_\star$, $T_\mathrm{hi}$,
$T_\mathrm{lo}$, $\rho_\mathrm{int}$, $H_\star$,
$\delta_\mathrm{atm}$, ${X_k}_\star$, ${X_k}_\mathrm{atm}$, and $g$.
We then integrate outwards from the base of the ramp region ($z =
H_\star$), enforcing the discrete form of hydrostatic equilibrium,
Eq.~\ref{eq:hse}.  Integrating upwards, we find $p_i$ and $\rho_i$
using a Newton-Raphson solver together with the equation of state,
with either the temperature specified, $T_i = T(p_i, \rho_i,
\{X_k\}_i)$ (in the isothermal, ramp, and buffer layers) or constant
entropy, $s_i = s(p_i, \rho_i, \{X_k\}_i)$, in the fuel layer.  This
follows the procedures described in \citet{ppm-hse}.  We use a
constant temperature for all $z < H_\star + 3\delta_\mathrm{atm}$.
Above this, we switch to an isentropic atmosphere until the
temperature drops to \deleted{a floor value, }$T_\mathrm{lo}$, at
which point we again keep the temperature constant.  The integration
of the atmosphere continues until the density falls to a low density
cutoff, $\rho_\mathrm{cutoff}$.  The material above this height is
taken to have constant density and temperature.  The choice of factors
in front of $\delta_\mathrm{atm}$ were designed to make sure the peak
$T$ is attained at the desired density of the burning layer.  The
parameters we use for the model generation are listed in
Table~\ref{table:params} and the initial model profiles are showing in
Figure~\ref{fig:initial_models}.

\begin{deluxetable}{lcc}
\tablecaption{\label{table:params} Initial model parameters.}
\tablehead{\colhead{parameter} & \colhead{cool} & \colhead{hot}}
\startdata
$T_\star$       & \multicolumn{2}{c}{$10^8$~K} \\
$T_\mathrm{hi}$ & $2\times 10^8$~K & $1.4\times 10^9$~K \\
$T_\mathrm{lo}$ & \multicolumn{2}{c}{$8\times 10^6$~K} \\
$\rho_\mathrm{int}$ & \multicolumn{2}{c}{$3.43\times 10^6~\gcc$} \\
$\rho_\mathrm{fuel}$\tablenotemark{a} & $2.36\times 10^6~\gcc$ & $1.20\times 10^6~\gcc$ \\
$\rho_\mathrm{cutoff}$ & \multicolumn{2}{c}{$10^{-4}~\gcc$} \\
$H_\star$       & \multicolumn{2}{c}{2000~cm} \\
$\delta_\mathrm{atm}$ & \multicolumn{2}{c}{50~cm} \\
$X_\star(\isotm{Ni}{56})\tablenotemark{b}$ & \multicolumn{2}{c}{1.0} \\
$X_\mathrm{atm}(\isotm{He}{4})\tablenotemark{b}$ & \multicolumn{2}{c}{1.0} \\
$\gb$             & \multicolumn{2}{c}{$-1.5\times 10^{14}~\mathrm{cm~s^{-2}} \hat{\bf z}$} \\
\enddata
%
\tablenotetext{a}{This is not an input parameter, but instead is
  computed during integration.  We list it here for reference.}
%
\tablenotetext{b}{All other species are taken as 0.}
\end{deluxetable}

We blend the hot and cold models laterally to produce the perturbation
needed to initiate a localized flame, with the hot model at the
origin of the axisymmetric geometry.  The blending is done as:
\begin{align}
p(r,z) &= f(r) p_\mathrm{hot}(z) + [1-f(r)] p_\mathrm{cool}(z) \\
\rho(r,z) &= f(r) \rho_\mathrm{hot}(z) + [1-f(r)] \rho_\mathrm{cool}(z) \\
X_k(r,z) &= f(r) {X_k}_\mathrm{hot}(z) + [1-f(r)] {X_k}_\mathrm{cool}(z)
\end{align}
with
\begin{equation}
f(r) = \begin{cases}
     1 & r < r_\mathrm{pert} \\
   1 - \frac{r - r_\mathrm{pert}}{\delta_\mathrm{blend}} & r_\mathrm{pert} \le r \le r_\mathrm{pert} + \delta_\mathrm{blend} \\
     0 & r > r_\mathrm{pert} + \delta_\mathrm{blend}
\end{cases}
\end{equation}
Since the equation of hydrostatic equilibrium is linear and our
blending is a linear combination of two models in hydrostatic
equilibrum, the blended model is also in vertical equilibrium initially.  We choose
$r_\mathrm{pert} = 1.024\times 10^4$~cm and $\delta_\mathrm{blend} = 2048$~cm.  Once
the blended model is constructed, we compute $T(r,z)$ and $(\rho e)(r,z)$
from the equation of state,
\begin{align}
  T(r,z) &= T(\rho(r,z), p(r,z), X_k(r,z)) \\
  (\rho e)(r,z) &= \rho(r, z) \cdot e(\rho(r,z), p(r,z), X_k(r,z)) 
\end{align}
